A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-12 main.tex
Intro WIP
Lemma 29 counterexample move to appendix

This paper presents the first full mechanical formalization of the metatheory for
higher-ranked polymorphic type inference. The system that we formalize
is the bidirectional type system by Dunfield and
Krishnaswami (DK). The DK type system has two variants (a declarative
and an algorithmic one) that have been \emph{manually} proven
\emph{sound}, \emph{complete} and \emph{decidable}. We present 
a mechanical formalization in the Abella theorem prover 
of DK's declarative type system with a novel algorithmic system. We have
several reasons to use a new algorithm. Firstly, our new algorithm
employs \emph{worklist judgements}, which allow precise capture of the
scope of variables and simplify the formalization of scoping in a
theorem prover. Secondly, while DK's original formalization comes with very
well-written manual proofs, there are several details missing,
which complicate the task of writing a formal proof. 
Perhaps more importantly, some proofs/lemmas are wrong and it is
not clear how to replace these by other valid proofs/lemmas. 
\bruno{others?} Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. Since such type inference algorithms are quite
subtle and have a complex metatheory, we believe that mechanical
formalizations are an important advance in type-inference research.
\end{abstract}

@@

Text of appendix \ldots

>>>>

This paper presents the first full mechanical formalization of the metatheory for
higher-ranked polymorphic type inference. The system that we formalize
is the bidirectional type system by Dunfield and
Krishnaswami (DK). The DK type system has two variants (a declarative
and an algorithmic one) that have been \emph{manually} proven
\emph{sound}, \emph{complete} and \emph{decidable}. We present 
a mechanical formalization in the Abella theorem prover 
of DK's declarative type system with a novel algorithmic system. We have
several reasons to use a new algorithm. Firstly, our new algorithm
employs \emph{worklist judgements}, which allow precise capture of the
scope of variables and simplify the formalization of scoping in a
theorem prover. Secondly, while DK's original formalization comes with very
well-written manual proofs, there are several details missing,
which complicate the task of writing a formal proof. 
Moreover, some proofs/lemmas are wrong and it is
not clear how to replace these by other valid ones. 
\bruno{others?} Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. Since such type inference algorithms are quite
subtle and have a complex metatheory, we believe that mechanical
formalizations are an important advance in type-inference research.
\end{abstract}

@@

\jimmy{Lemma 29 is false. TODO structure and more text}
The first conclusion of the lemma says that if $\Gm_L\longrightarrow \Delta_R$ and
$\Gm_L,\Gm_R\longrightarrow \Delta_L,\Delta_R$ then
$\Gm_L,\al,\Gm_R\longrightarrow \Delta_L,\al,\Delta_R$.
If we pick $\Gm_L = \nil, \Gm_R = \bt, \Delta_L = \bt, \Delta_R = \nil$, the conditions
$\nil\longrightarrow \bt$ and $\bt\longrightarrow \bt$ hold, but the conclusion
$\al,\bt \longrightarrow \bt,\al$ does not hold.

### 2019-02-12 sections/Overview.tex
quick notes
Lemma 29 counterexample move to appendix

\paragraph{Overlapping rules}
Some of the declarative rules overlaps with each other.
Declarative subtyping rules $\mathtt{{\le}\forall L}$ and $\mathtt{{\le}\forall R}$
both match the conclusion $\Psi\vdash \forall a. A \le \forall a. B$.
In such case, choose $\mathtt{{\le}\forall R}$ first is always better,
since we introduce the type variable $a$ to the context earlier,
which gives more flexibility on the choice of $\tau$.

@@

However, Lemma 29 (Parallel Admissibility) is false.
The first conclusion of the lemma says that if $\Gm_L\longrightarrow \Delta_R$ and
$\Gm_L,\Gm_R\longrightarrow \Delta_L,\Delta_R$ then
$\Gm_L,\al,\Gm_R\longrightarrow \Delta_L,\al,\Delta_R$.
If we pick $\Gm_L = \nil, \Gm_R = \bt, \Delta_L = \bt, \Delta_R = \nil$, the conditions
$\nil\longrightarrow \bt$ and $\bt\longrightarrow \bt$ hold, but the conclusion
$\al,\bt \longrightarrow \bt,\al$ does not hold.

@@

>>>>

\paragraph{Overlapping rules}
\jimmy{Move to the Alg. Section}
Some of the declarative rules overlaps with each other.
Declarative subtyping rules $\mathtt{{\le}\forall L}$ and $\mathtt{{\le}\forall R}$
both match the conclusion $\Psi\vdash \forall a. A \le \forall a. B$.
In such case, choose $\mathtt{{\le}\forall R}$ first is always better,
since we introduce the type variable $a$ to the context earlier,
which gives more flexibility on the choice of $\tau$.

@@

Moreover, Lemma 29 (Parallel Admissibility) is false.
One may find a counterexample with detailed explaination in the appendix.
Lemma 29 is an important lemma for the proof of soundness and completeness,
which are two main results of their metatheory.
Instantiation soundness (i.e. a part of subtyping soundness) and typing completeness
both directly require Lemma 29.
Lemma 54 (Typing Extension) requires Lemma 29 and is applied 13 times
in the proof of typing soundness and completeness.

@@

\jimmy{Compare with DK's algo: subsumption rule might cause leak}
